### CVEN 6833 - Final Project
#### Zachary Carpenter
#### Spring 2025

This R Markdown file runs the main body of code for this final project analysis.

##### *Get Libraries*
```{r Get Libraries, warning = FALSE, message = FALSE}
library(dplyr)
library(lubridate)
library(maps)
library(akima)
library(fields)
library(kohonen)
library(ggplot2)
library(tidyverse)
library(tibble)
library(randomForest)
library(patchwork)

```

##### *Source Functions*
Each of these function codes can be found at the end of this document as "commented out" lines of code. The source commands here pull from the same directory/repository as this Rmd file.
```{r Source Functions}
source("RFMod4FLow.R")
source("modeval.R")

```

##### *Read and Format Flow Data*
```{r Read Flow Data, warning=FALSE}
flowdata_raw <- as.data.frame(read.csv("NatFlowDataCRB_1906_to_2020.csv", header = TRUE, skip = 3))
flowdata_raw <- flowdata_raw[1:(nrow(flowdata_raw)-5),]   # Remove summary rows

flowdata_clean <- flowdata_raw[, !sapply(flowdata_raw, anyNA)]    # Remove NA columns

flowdata_ub <- flowdata_clean[, 1:21]   # Restrict to just upper basin gages

flowdata_ub$Date <- ymd(flowdata_ub$Date)    # Converts date column to date type object

flowdata_ub_AMJ <- flowdata_ub %>% filter(month(Date) %in% 4:6)   # Filter to just spring flow (A-J)

# Convert flow data to numeric and remove commas from entries
flowdata_ub_AMJ[, 2:ncol(flowdata_ub_AMJ)] <- lapply(
  flowdata_ub_AMJ[, 2:ncol(flowdata_ub_AMJ)], 
  function(col) as.numeric(gsub(",", "", col)))


# Loop structure to sum the annual spring flow for each year and Upper Basin gage location
ann_spring_flow <- matrix(nrow = length(1906:2020), ncol = ncol(flowdata_ub_AMJ)-1) 
yrs <- as.numeric(1906:2020)

for(i in 1:length(yrs)){
  for(j in 2:ncol(flowdata_ub_AMJ)){
    ann_spring_flow[i, j - 1] = sum(flowdata_ub_AMJ[[j]][year(flowdata_ub_AMJ$Date) == yrs[i]], na.rm = TRUE)}}

ann_spring_flow <- ann_spring_flow/(10^6)   # Convert to MAF

# Add gage names and years to the annual spring flow data, convert to data frame
rownames(ann_spring_flow) <- yrs
colnames(ann_spring_flow) <- colnames(flowdata_ub_AMJ[2:ncol(flowdata_ub_AMJ)])
ann_spring_flow <- as.data.frame(ann_spring_flow)

```

##### *Cmds-4Zach.R*
This code chunk was directly provided/written by Professor R. Balaji.
```{r Cmds-4Zach.R, warning=FALSE}
nrows=72
ncols=20
ntime = 115    #Dec-Feb 1906 - Dec-Feb 2020

nyrs = ntime    #1906 - 2020

nglobe = nrows*ncols
N = nrows*ncols

### Lat - Long grid..
ygrid=seq(-22.5,72.5,by=5)
ny=length(ygrid)

xgrid=seq(27.5,382.5,by=5)
#xgrid[xgrid > 180]=xgrid[xgrid > 180]-360	#longitude on 0-360 grid if needed
nx=length(xgrid)

xygrid=matrix(0,nrow=nx*ny,ncol=2)

i=0
for(iy in 1:ny){
for(ix in 1:nx){
i=i+1
xygrid[i,1]=ygrid[iy]
xygrid[i,2]=xgrid[ix]
}}

# Read Kaplan SST data..
data=readBin("https://civil.colorado.edu/%7Ebalajir/CVEN6833/HWs/HW-2/data.r4", what="numeric", n=(nrows * ncols * ntime), size=4,endian="swap")

data <- array(data = data, dim=c( nrows, ncols, ntime ) )

data1=data[,,1]

# the lat -long data grid..
index=1:(nx*ny)

index1=index[data1 < 20 & data1 != "NaN"]	# only non-missing data.
xygrid1=xygrid[index1,]

x1=xygrid1[,2]
#x1[x1 < 0]= x1[x1 < 0] + 360
#xygrid1[,2]=x1

nsites=length(index1)
data2=data1[index1]

### SSTdata matrix - rows are seasonal (i.e. one value per year)
## and columns are locations
sstdata=matrix(NA,nrow=nyrs, ncol=nsites)

for(i in 1:nyrs){
data1=data[,,i]
index1=index[data1 < 20 & data1 != "NaN"]
data2=data1[index1]
sstdata[i,]=data2
}

sstannavg = sstdata
indexgrid = index1
rm("data")	#remove the object data to clear up space

## write out the grid locations..
write(t(xygrid1),file="kaplan-sst-locs-4Zach.txt",ncol=2)

```

##### *SOM on Winter SST*
```{r SOM on Winter SST, warning=FALSE}
# Establish the SOM Parameters
n = 3

nnodes = n^2

# Perform the SOM Computations
library(kohonen)
begin <- Sys.time()
SOM <- som(sstannavg, grid = somgrid(n,n, "rectangular"))
end <- Sys.time()
end-begin

# Plot the number of observations in each node
colors <- function(n, alpha = 1){
  rev(heat.colors(n, alpha))}

plot(SOM, type = "counts",
     palette.name = colors,
     heatkey = TRUE, main = "Number of Observations per Node")

# Write the composite data nodes
nodes = as.data.frame(array(NaN, dim = c(dim(sstannavg)[2],nnodes)))
for(i in 1:nnodes){
  index = which(SOM$unit.classif == i)
  sst_df = sstannavg[index,]
  if(length(index)>1){
    nodes[,i] = colMeans(sst_df, na.rm=T)}
  else{
    nodes[,i] = sst_df}}

noderange = range(nodes)

# Plot the SST SOM Nodes - both in a 3x3 grid
lon = sort(unique(xygrid[,2]))
lat = sort(unique(xygrid[,1]))

sstrange = range(nodes)

nx = 50

par(oma = c(0,0,2.5,1))
par(mfrow = c(n, n))
par(mar = c(2, 3, 2, 1))

detach(package:kohonen)

for(i in 1:nnodes){
  zfull = rep(NaN,nglobe)
  zfull[indexgrid] = nodes[,i]
  zmat = matrix(zfull, nrow = nrows, ncol = ncols)
  image.plot(lon, lat, zmat, ylim = range(-60, 90), main = paste("Node ", i, sep = ""))
  contour(lon, lat, (zmat), ylim = range(-60, 90), add = TRUE, nlev = 6, lwd = 1)
  maps::map("world2", add = T)}

mtext("Annual Average Winter SST - SOM Maps 3x3", outer = TRUE, side = 3, cex = 1.2, line = 1)

# Get the data indices for each node
obs_nodes <- SOM$unit.classif
node_idx <- split(1:nrow(sstannavg), obs_nodes)

```

##### *Select Gages and Boxplot the Flows for Each SOM Node*
The selected gages are the Upper Colorado Reach Inflow (Colorado River @ Glenwood Springs, CO), Gunnison River @ Grand Junction, CO, Green River Below Fontenelle Reservoir, WY, Little Snake River near Lily, CO, San Juan River near Bluff, UT, and Colorado River @ Lees Ferry, AZ.    
```{r Select Gage and Boxplot, warning=FALSE}
# Subset the annual spring flow dataset for just the selected gages
selectgage <- c(1, 6, 9, 13, 19, 20)

gagenames <- c("CR-Glenwood", "Gunnison", "Green", "Little Snake River", "San Juan River", "CR-LeesFerry")

ngages <-length(selectgage)

ann_spring_flow_sel <- ann_spring_flow[, selectgage]
colnames(ann_spring_flow_sel) <- gagenames

# Parse the spring flow years at the selected gages into their respective SOM nodes
nodeflows <- list()
node_count <- 1

for (nodenum in 1:nnodes){
  nodeflows[[nodenum]] <- ann_spring_flow_sel[node_idx[[nodenum]],]
  node_count <- node_count + 1
  }

# Boxplot the associated annual gage flows for each SOM node
par(mfrow = c(3,3), oma = c(0, 0, 3, 0))
for (nodenum in 1:nnodes){
  boxplot(as.matrix(nodeflows[[nodenum]][, 1:ngages]), ylim = c(-0.5,2), col = "lightblue", 
          ylab = "Flow (MAF)", main = paste("Node", nodenum), names = FALSE, cex.lab = 0.9, cex.main = 0.9)
  grid(nx = NA, ny = NULL)
}
mtext("Spring Flow for Each Gage at the 3x3 SOM Nodes", outer = TRUE, side = 3, cex = 0.9, line = 1)

par(mfrow = c(1,1), oma = c(3, 0, 0, 0))
for (nodenum in 1:nnodes){
  boxplot(as.matrix(nodeflows[[nodenum]][, 1:ngages]), ylim = c(-0.5,2), col = "lightblue", 
          ylab = "Spring Flow (MAF)", main = paste("Node", nodenum), names = gagenames, las = 2, cex.axis = 0.7)
  grid(nx = NA, ny = NULL)
}

```

##### *PCA on SSTs*
```{r PCA on SSTs, warning=FALSE}
# Perform the PCA
sst_pcaresult <- prcomp(sstannavg, scale. = TRUE)

# Extract the first 4 PCs
sstpcs <- as.data.frame(sst_pcaresult$x[,1:4])

```

##### *PCA on Spring Flows*
```{r PCA on Flows, warning=FALSE}
# Perform the PCA
flow_pcaresult <- prcomp(ann_spring_flow_sel, scale. = TRUE)

# Extract the first 2 PCs
flowpcs <- as.data.frame(flow_pcaresult$x[,1:2])

```

##### *Fit a Random Forest Model and predict for Spring flow PCs*
```{r RF Model, warning=FALSE}
# Fit a Random Forest Model for each of the Flow PCs and get predicted values
RF_mods <- list()

for (i in 1: ncol(flowpcs)){
  RF_mods[[i]] <- RFMod4Flow(flowpcs[,i], sstpcs)
}

flow_PC1_pred <- RF_mods[[1]]$preds
flow_PC2_pred <- RF_mods[[2]]$preds

flow_PCs_pred <- cbind(flow_PC1_pred, flow_PC2_pred)

```

##### *Predict spring flow at each gage using the RF Models*
```{r Predict with RF, warning=FALSE}
# Predict stream flow PCs with the RF models and then determine the scaled and centered flow values at every gage
flow_PCs_pred_ALL <- cbind(flow_PCs_pred, matrix(0, nrow(flow_PCs_pred), 4))

pred_flows_scaled <- flow_PCs_pred_ALL %*% t(flow_pcaresult$rotation)

# Unscale and uncenter the data to get the actual flow values for the gages
pred_flows_unscale <- sweep(pred_flows_scaled, 2, flow_pcaresult$scale, "*")
pred_flows <- sweep(pred_flows_unscale, 2, flow_pcaresult$center, "+")

```

##### *Evaluate the predictions*
```{r Model Evaluation, warning=FALSE}
# Evaluate each model fit using Correlation and RMSE
actvals <- as.data.frame(ann_spring_flow_sel)
predvals <- as.data.frame(pred_flows)

mod_eval <- modeval(actvals, predvals)

evalplots <- mod_eval$corrplot + mod_eval$rmseplot +
  plot_annotation(title = paste("Correlation and RMSE of the RF Predictions for All Gages"))
print(evalplots)

# Plot the predicted values against the actual values
predvact_plots <- list()

for (i in 1:ngages) {
  # Create a data frame for ggplot
  plot_data <- data.frame(Actual = actvals[, i], Predicted = predvals[, i])
  
  # Generate the plot object
  predvact_plots[[i]] <- ggplot(plot_data, aes(x = Actual, y = Predicted)) +
    geom_point() + 
    geom_abline(slope = 1, intercept = 0, color = "red", linetype = "solid") +
    labs(title = gagenames[i], x = "Actual", y = "Predicted") +
    theme(plot.title = element_text(size = 13, face = "bold", hjust = 0.5))
}
# Print the plots
predvact_plot_print <- predvact_plots[[1]] + predvact_plots[[2]] + predvact_plots[[3]] + predvact_plots[[4]] + predvact_plots[[5]] + predvact_plots[[6]] + 
  plot_annotation(title = "Predicted vs. Actual Spring Flow (MAF) at Each Gage",
  theme = theme(plot.title = element_text(hjust = 0.5, size = 14, face = "bold")))

print(predvact_plot_print)


```

##### *Commented Out Functions*
```{r Functions, warning=FALSE}
### RFMod4FLow.R
# RFMod4Flow <- function(flowpcs, sstpcs)
# {
#   # Fit a Random Forest Model
#   rf_mod <- randomForest(x = sstpcs, y = flowpcs, ntree = 1000, mtry = ncol(sstpcs),
#                                  importance = TRUE)
#   
#   # Predict off the Random Forest Model
#   pred_vals <- predict(rf_mod, newdata = sstpcs)
#   
#   return(list(mod = rf_mod, preds = pred_vals))
# }


### modeval.R
# modeval <- function(actvals, predvals){
# # Calculate Correlation and RMSE
# n <- ncol(actvals)
# corrskill <- numeric(n)
# rmseskill <- numeric(n)
# 
# for (i in 1:n){
#   corrskill[i] <- cor(actvals[,i], predvals[,i])
#   rmseskill[i] <- sqrt(mean((actvals[,i] - predvals[,i])^2))
# }
# 
# # Create data frames for plotting
# corr_df <- data.frame(metric = "Correlation", value = corrskill)
# rmse_df <- data.frame(metric = "RMSE", value = rmseskill)
#   
# # Boxplot the Correlation and RMSE
# corrplot <- ggplot(corr_df, aes(x = metric, y = value)) +
#   geom_boxplot() +
#   labs(title = "Corr.", x = "", y = "")+
#   theme(plot.title = element_text(size = 13, face = "bold", hjust = 0.5))
# 
# rmseplot <- ggplot(rmse_df, aes(x = metric, y = value)) +
#   geom_boxplot() +
#   labs(title = "RMSE", x = "", y = "")+
#   theme(plot.title = element_text(size = 13, face = "bold", hjust = 0.5))
# 
# return(list(corrplot = corrplot, rmseplot = rmseplot))
# }

```






